Solution of Differential Algebraic Equations with Manifold Constraints on the Latent Space of an Autoencoder

### 2019-06-09 paper/paper.tex
writing up edits

            pdftitle={Solution of Differential Algebraic Equations with Manifold Constraints on the Latent Space of an Autoencoder},

@@

\begin{document}
\maketitle
\begin{abstract}
We train an autoencoder to determine the lower-dimensional
representation of the EOS surface using modern deep learning techniques
without using any phase labels. The decoder end of the model is then
plugged back into the balance laws, eliminating the constraint equation.
\end{abstract}

@@

Accurately simulating multispecies reaction and transport with
undergoing phase transitions is a critical aspect to a variety of
problems.
Motivations for this work are subsurface energy systems such as shale
and oceanic natural gas reservoirs in shales or engineered geothermal
systems.

A single material with multiple phase transitions
is a challenge in of itself, and the descriptions become combinatorially
more complex as species and phases are added. This results in
combinatorially more human labor and combinatorially more code which
tends to involve complex branching logic. The descriptions then pose
numerical challenges when incorporated into macroscale simulations due
to obstacles such as discrete phase state machines and primary variable
switching.

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot flux_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot flux_{CO_2} + s \\
  \partial_t e& = \nabla \cdot flux_{heat} + s
\end{align}
More species add more rows, and applicaitons also involve continuum
mechanics and chemical reactions! The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}

This work only considers the "single beaker" problem; the flow
and transport aspects have similar challenges. (The next stage of this
work will "connect the beakers".)

In section 2, we will cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \leftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
We will use this set of possible values viewpoint to train an
autoencoder on the set of $u$ for a latent space parameterized by $q$:
\begin{equation}
u = D(q), \; q=E(u) \;\forall u \in U
\end{equation}
which satisifies
\begin{equation}
  D(q)\in U \forall q.
\end{equation}
We can then write the original equation by
\begin{equation}
z(q,\partial_t q)=0
\end{equation}

 \cite{wu_physics-informed_2018}
\cite{xie_tempogan:_2018}

@@

The actual meaning of the parameterization is not important, so long as
it can be effectively solved and then \emph{decoded} back to the
physically meaningful quantities.

We redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. Because the EOS was originally derived from data, a data-driven
approach lends itself in which the EOS is a set of points in
temperature-pressure-density-enthalpy space. (In the multicomponent
cases there are more dimensions.)

@@

@@

The notebooks and software implementations are released open source to
accompany this work. They can be found at the github repository
located at: \href{https://github.com/afqueiruga/latentsim}
Certain components are factored into submodule repositories which are
each released under the LGPL at seperate repositories referenced in
the implementation details.

\hypertarget{header-n3234}{%
\section{Multiphase Equations of State}\label{header-n3234}}

For this preliminary work, we consider only a single-beaker problem
without flow. Let \(\rho\) be the material density, \(T\) be the
temperature, \(p\) be the material pressure, \(h\) be the material
enthalpy, and \(u\) be the internal energy. There are two equations for
balance of mass and energy.

\begin{align}
\partial_t \rho & = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t \rho u & = \mathbf{k}_T(T_\infty-T) + s
\end{align}

The beaker is exposed to a reservoir with \(p_\infty\) and \(T_\infty\)
via a mixed boundary condition straw with transport and conduction
coefficients \(\mathbf{k}_p\), \(\mathbf{k}_T\). Mass and heating
sources and sinks are the terms \(r\) and \(s\).\footnote{We express the
  material by enthalpy instead of internal energy; they are related by
  the relation \(\rho u = \rho h - p\). Future work may work directly
  with internal energy.}

The typical methodology is to use the empirical relations for density
and enthalpy \(h\) as a function of pressure and temperature and solve
the DAE for \(p\) and \(T\) implicitly:

\begin{align}
\text{Solve for}\, p(t)\, \text{and}\, T(t)\, \text{such that:}\\
\partial_t \rho(p,T) & = \nabla \cdot \mathbf{k}\nabla p + r\\
\partial_t \rho u(p,T) & = \nabla \cdot \mathbf{k'}\nabla T + s
\end{align}

The complicatation is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.

We also consider the phase equilibria; at a larger scale it is possible
for a control volume to be a mixture of two phase coexisting, such as
boiling with some mixture saturation \(S_{gas}=1-S_{liq}\).
(For a very fine scale simulation, bubbles would be explicitly modeled.)
The material can exist on this section of the surface as a mixture of
the two phases.

\[\rho = X_{H_2O}\left(S_{gas} \rho_{gas} + S_{liq} \rho_{liq} + S_{ice} \rho_{ice} + S_{crit} \rho_{crit} \right)\]

One possible definition of phase boundaries:

\begin{quote}
Phase transitions occur when the thermodynamic free energy of a system
is non-analytic for some choice of thermodynamic variables (cf. phases).

https://en.wikipedia.org/wiki/Phase\_transition
\end{quote}

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/phase_diagram.png}
\caption{Phase boundaries for water }
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/water_eos.png}
\caption{\label{fig:pTrho} The $\log{p}$, $T$, $\rho$ equation of
  state surface for water. This plot spans four phases, three
  equilibria, and the triple point, which are each given their own color.
  There is a fourth dimension to the surface of enthalpy $h$ that exhibits
  similar discontinuities; the $\log{p}$, $T$, $h$ surface is plotted
  in the Appendix.}
\end{figure}

There are even more known phases of crystal ice which for water
For water, the supercritical phase and exotic ices only occur in
laboratory conditions or naturally near the
crust \cite{}. For other materials these exotic phases can be
applicable. For example, in geoengineered systems using $CO_2$
such as geothermal plants are carbon sequestration facilities, the
working fluid is usually in the supercritical regime at depth.

@@

Formulate balance equations as

\[R(X ; phase) = 0\]

Solve the differentiable part:
\[\frac{\partial R}{\partial X}\Delta X^{k+1/2} = -R(X^k ; phase^k)\]
and iterate on the nondifferential part:
\[phase^{k+1},X^{k+1} = statemachine(X^{k+1/2} ; phase^k)\]
The phase switching logic has the form of a switch case:

@@

We instead replace our problem with two balance laws and one unknown
constraint:

\begin{align}
\text{Solve for}\, \rho(t), \, h(t), \, p(t),\, \text{and}\, T(t)\, \text{satisfying:}\\
\partial_t \rho &  = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t (\rho h-p) & = \mathbf{k}_T(T_\infty-T) + s \\
\text{such that they lie on the material EOS,}\\
\rho,h,p,T & \in \{\text{EOS}\}
\end{align}

The \emph{ground truth} is the experimental data in the first place; the
branching curve fit is only one realization of representing the data. We
have demonstrated that it is not possible to represent any two of these
in terms of the other two; and even if so, it is not necessarily a good
and robust fit. What if we search for a new fit?

@@

An archetypical constraint problem the pendulum. Posed as dynamics
problem with a Lagrange-multipler-esque centripetal force, it
has three unknowns wtih two second-order equations and a constraint,

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
x^2 + y^2 & = R^2
\end{align}

Note that we are considering the nonlinear pendulum.

@@

The manifold for this simple problem is smooth, so two nested
polynomials forms the autoencoder:
\[\left\{\begin{array}{c} x\\y\end{array}\right\}
\rightarrow W_{enc} x^n \rightarrow q \rightarrow W_{dec} x^n \rightarrow 
\left\{\begin{array}{c} x\\y\end{array}\right\}\]

(We explicitly write the bias \(+b\); it is equivalent to including
\(x^0=1\) in the polynomial set.)
The equations
of motion given some parameter space that satisfies the constraint is

\begin{equation}
\frac{\mathrm{d}q}{\mathrm{d}t}\frac{\partial L}{\partial \dot{q}} = \frac{\partial L}{\partial q}.
\end{equation}

We'll use the autoencoder to generate the parameterization of \(x(q)\)
and \(y(q)\). By the chain rule, the velocities are related to the
feature space by
\begin{equation}
\dot{x} = \frac{\mathrm{d}x(q)}{\mathrm{d}t} = \frac{\partial x}{\partial q}\frac{\mathrm{d}q}{\mathrm{d}t}
\end{equation}
and similarly for the \(y\) direction.

When we want to solve the dynamics, we can treat the autoencoder the
same as \(x(\theta)\), plugging in
\[L(x(q),v(q,\dot{q}))\]
Taking the partial derivatives with respect to the feature variable and
its rate, the components to the equation of motion are build the DAE we know from physics:

\[\frac{\mathrm{d}}{\mathrm{d}t}\frac{\partial L}{\partial \dot{q}} = \frac{\partial L}{\partial q}\]
where we get the velocity with the chain rule:
\[v = \frac{\partial x}{\partial q} \dot{q}\]

Then plug it into a Runge Kutta (trapezoidal) to solve with Newton's
method:
\[\frac{\partial L}{\partial \dot{q}}_i - \Delta t \frac{\partial L}{\partial q}_i = \frac{\partial L}{\partial \dot{q}}_0 + \Delta t \frac{\partial L}{\partial q}_0\]

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/pendulum_q.png}
\caption{}
\end{figure}

@@

\bibliographystyle{plainnat}
\bibliography{zotero}

>>>>

            pdftitle={Solution of Differential Algebraic Equations with Local Constraints on the Latent Space of an Autoencoder},

@@

\begin{document}
\maketitle
\begin{abstract}
Multiphase simulation is hard.

We train an autoencoder to determine the lower-dimensional
representation of the EOS surface using modern deep learning techniques
without using any phase labels. The decoder end of the model is then
plugged back into the balance laws, eliminating the constraint equation.
\end{abstract}

@@

Accurately simulating multispecies reaction and transport with
undergoing phase transitions is a critical aspect to a variety of
problems.
This work is motivated by advancing simulation techniques for subsurface energy systems such as shale
and oceanic natural gas reservoirs in shales or engineered geothermal
systems.

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e& = \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}
(In this paper, we use use $\partial_t$, $\mathrm{d}/\mathrm{d}t$,
and $\dot{()}$ to denote differentiation with respect to time
interchangably depending on context.
The complixity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

% {\bf RESERVOIR PHYSICS LIT REVIEW HERE; Other applications;
%   plasticity, }

A single material with multiple phase transitions
is a challenge in of itself, and the descriptions become combinatorially
more complex as species and phases are added. This results in
combinatorially more human labor and combinatorially more code which
tends to involve complex branching logic. The descriptions then pose
numerical challenges when incorporated into macroscale simulations due
to obstacles such as discrete phase state machines and primary variable
switching.
This work only considers the "single beaker" problem; the flow
and transport aspects have similar challenges. (The next stage of this
work will "connect the beakers".)

We redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. We can cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
where $u=\{T,p,\rho,h\}$. We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \leftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
This maps directly to the original experimental procedure for
determining a constraint equation.

Because the EOS was originally derived from data, a data-driven
approach lends itself in which the EOS is a set of points in
temperature-pressure-density-enthalpy space. (In the multicomponent
cases there are more dimensions.)
We will use this set of possible values viewpoint to train an
autoencoder on the set of $u$ for a latent space parameterized by $q$:
\begin{equation}
u = D(q), \; q=E(u) \;\forall u \in U
\end{equation}
which satisifies
\begin{equation}
  D(q)\in U \forall q.
\end{equation}
We can then write the original equation by
\begin{equation}
z(q,\partial_t q)=0
\end{equation}
The actual meaning of the parameterization is not important, so long as
it can be effectively solved and then \emph{decoded} back to the
physically meaningful quantities.

% {\bf ML LIT REVIEW HERE} % TODO
\cite{wu_physics-informed_2018}
\cite{xie_tempogan:_2018}

 A program synthesis approach was used to search for and evaluate possible methods of Peridynamics in Queiruga, 2017 \cite{queiruga2017numerical}. (An open source implementation of the study can be found Periflakes)\cite{queiruga2018periflakes}.

@@

@@

We build a test dataset off of the IAPWS solutions. The proposed
methodology applies to experimental data directly, or preexisting
compendiums of empirical fits and theoretical relations.
Note that, even if a relation can be theoretical derived, that does
not mean that the analytical relationship yields the best program
implementation: it is often the case that numerial approximations are
superior to closed-form mathematical solutions in computational
efficiency, accuracy, and numerical stability.

@@

The notebooks and software implementations are released open source to
accompany this work. They can be found at the github repository
located at: \href{https://github.com/afqueiruga/latentsim}.
Certain components are factored into submodule repositories which are
each released under the LGPL at seperate repositories referenced in
the implementation details.

\section{Multiphase Equations of State}

% \begin{figure}
%    CSTR
%   \caption{ \label{fig:cstr} Illustration of the Continuous stirred-tank
%   reactor. The material in the reactor is exposed to a reservoir with pressure
%   $p_\inf$ and temperature $T_\inf$ via a conduit with permeability
%   and thermal conductivities $k_p$ and $k_T$, direct forced mass
%   injection/withdrawl $r$, or heating and cooling $s$.}
% \end{figure}

For this preliminary work, we consider only a single-beaker problem
without flow, a continuous stirred-tank reactor (CSTR), as illustrated
in Figure \ref{fig:cstr}. Let \(\rho\) be the material density, \(T\) be the
temperature, \(p\) be the material pressure, \(h\) be the material
enthalpy, and \(u\) be the internal energy. There are two equations for
balance of mass and energy.
\begin{align}
\partial_t \rho & = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t \rho u & = \mathbf{k}_T(T_\infty-T) + s 
\end{align}
The beaker is exposed to a reservoir with \(p_\infty\) and \(T_\infty\)
via a mixed boundary condition straw with transport and conduction
coefficients \(\mathbf{k}_p\), \(\mathbf{k}_T\). Mass and heating
sources and sinks are the terms \(r\) and \(s\).\footnote{We express the
  material by enthalpy instead of internal energy; they are related by
  the relation \(\rho u = \rho h - p\). Future work may work directly
  with internal energy.}
We are neglecting upstreaming in the connection between the reservoir
and the container.

We also consider the phase equilibria: beyond the scale of resolving
individual bubbles, it is possible
for a control volume to be a mixture of two phase coexisting.
(For a very fine scale simulation, e.g. at the pore scale, bubbles
would be explicitly modeled.)
We introduce a saturation $S_{phase}\in[0,1]$ for each phase;.
The material can exist on this section of the surface as a mixture of
the two phases. Thus, its density is calculated by a mixing the
properties of all phases at that point in the phase diagram:
\begin{equation}
  \rho = S_{gas} \rho_{gas} + S_{liq} \rho_{liq} + S_{ice}
    \rho_{ice} + S_{crit} \rho_{crit}
  \end{equation}
Each phase
and equilibrium surface constrains the possible saturations. For example, on the boiling equilibrium surface there are two phases with some mixture
saturations \(S_{gas}=1-S_{liq}\) and $S_{ice}=S_{crit}=0$. 
For water, three phases can exist in equilibrium. Given more species, it is possible for an arbitrary number of phases
to be in equilibrium; there is a quadruple point in methane-water
hydrates.

But, what is a phase boundary? The traditional explanation is shown in
Figure \ref{fig:phaseboundaries}. The use of intensive properties breaks down:\footnote{https://en.wikipedia.org/wiki/Phase\_transition}

\begin{quote}
Phase transitions occur when the thermodynamic free energy of a system
is non-analytic for some choice of thermodynamic variables (cf. phases).
\end{quote}

We can also assemble the surface of possible $T,p,\rho$ coordinates,
shown in Figure \ref{fig:pTrho}.
\begin{figure}
\centering
\includegraphics[width=4in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/phase_diagram.png}
\caption{\label{fig:phaseboundaries}Phase boundaries for water on the $T-p$ diagram. The standard
interpretation is that the properties of the material change
drastically when crossing this boundary. Note how it is possible to form a path from a
definitively gas state to a definitively liquid state without crossing
a phase boundary. The supercritical regime has no clear boundaries,
but requires a seperate curve fit in practice due to numerical challenges.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=4in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/water_eos.png}
\caption{\label{fig:pTrho} The $\log{p}$, $T$, $\rho$ equation of
  state surface for water in units of $\log{\text{Pa}}$, K, and \si{}{kg/m$^2$} This plot spans four phases, three
  equilibria, and the triple point, which are each given their own color.
  There is a fourth dimension to the surface of enthalpy $h$ that exhibits
  similar discontinuities; the $\log{p}$, $T$, $h$ surface is plotted
  in the Appendix. (Every point on this plot is a point in the
  generated dataset used for the optimization)}
\end{figure}

{\bf Remark:}There are even more known phases of crystal ice which for water
For water, the supercritical phase and exotic ices only occur in
laboratory conditions or naturally near the
crust \cite{}. For our case, we only handle the Ice I and cutoff the
regime at \si{10^8}{Pa}. For other materials these exotic phases can be
applicable. For example, in geoengineered systems using $CO_2$
such as geothermal plants are carbon sequestration facilities, the
working fluid is usually in the supercritical regime at depth. cite
curt. cite hydrates.
When binary mixtures are involved, the number of possible phases and
equilibria increases again.

{\bf Remark:} This interpretation breaks down with the existenceof super-saturated phases in non-stable regions: e.g., it is possible to
heat water beyond it's boiling point in a microwave. We are concerned
with environments where this would not happen (the very rough and
heterogenous) pores of rocks, so we have not thought of how to
integrate this problem.

@@

The typical methodology is to use the empirical relations for density
and enthalpy \(h\) as a function of pressure and temperature and solve
the DAE for \(p\) and \(T\) implicitly:
\begin{align}
\text{Solve for}\, p(t)\, \text{and}\, T(t)\, \text{such that:}\\
\partial_t \rho(p,T) & = \nabla \cdot \mathbf{k}\nabla p + r\\
\partial_t \rho u(p,T) & = \nabla \cdot \mathbf{k'}\nabla T + s
\end{align}
The complicatation is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.

\begin{tabular}{l|c|c}
  State & $X_1$ & $X_2$ \\
  \hline
  Ice & $p$ & $T$ \\
  Liquid & $p$ & $T$ \\
  Gas & $p$ & $T$ \\
  Supercritical & $\rho$ & $T$ \\
  Ice-Liquid & $S_{ice}$ & $T$ \\
  Ice-Gas & $S_{ice}$ & $T$ \\
  Gas-Liquid & $S_{gas}$ & $T$ \\
  Ice-Gas-Liquid & $S_{ice}$ & $S_{gas}$ 
\end{tabular}

  This requires a statemachine. For the complete states of water, it is shown in Figure \ref{fig:statemachine}.
  \begin{figure}
    \includegraphics[width=4in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/slides/phase_4.png}
    \caption{\label{fig:statemachine}States and possible transitions for water.}
\end{figure}
Formulate balance equations as
\begin{equation}
R(X ; phase) = 0
\end{equation}
Solve the differentiable part:
\[\frac{\partial R}{\partial X}\Delta X^{k+1/2} = -R(X^k ; phase^k)\]
and iterate on the nondifferential part:
\[phase^{k+1},X^{k+1} = statemachine(X^{k+1/2} ; phase^k)\]
The phase switching logic has the form of a switch case:

@@

We instead replace our problem with two balance laws and one unknown
constraint:
\begin{align}
\text{Solve for}\, \rho(t), \, h(t), \, p(t),\, \text{and}\, T(t)\, \text{satisfying:}\\
\partial_t \rho &  = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t (\rho h-p) & = \mathbf{k}_T(T_\infty-T) + s \\
\text{such that they lie on the material EOS,}\\
\rho,h,p,T & \in \{\text{EOS}\}
\end{align}
The \emph{ground truth} is the experimental data in the first place; the
branching curve fit is only one realization of representing the data. We
have demonstrated that it is not possible to represent any two of these
in terms of the other two; and even if so, it is not necessarily a good
and robust fit. What if we search for a new fit?

@@

An archetypical constraint problem is the pendulum. Posed as dynamics
problem with a Lagrange-multipler-esque centripetal force, it
has three unknowns wtih two second-order equations and a constraint,

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
x^2 + y^2 & = R^2
\end{align}
Note that we are considering the nonlinear pendulum.

@@

The manifold for this simple problem is smooth, so two nested
polynomials forms the autoencoder:
\[\left\{\begin{array}{c} x\\y\end{array}\right\}
\rightarrow W_{enc} x^n +b_{enc}\rightarrow q \rightarrow W_{dec} x^n +b_{dec}\rightarrow 
\left\{\begin{array}{c} x\\y\end{array}\right\}\]
where $x^n$ denotes a polynomial basis set, e.g. for two variables $x^n=\{x,y,x^2,xy,y^2,..y^n\}$.
(We explicitly write the bias \(+b\); it is equivalent to including
\(x^0=1\) in the polynomial set.)

We'll use the autoencoder to generate the parameterization of \(x(q)\)
and \(y(q)\). To do this, we optimize the loss function,
\begin{equation}
L(x) = \left\| x-D(E(x)) \right\|
\end{equation}

When we want to solve the dynamics, we can treat the autoencoder the
same as \(x(\theta)\), plugging in
\[L(x(q),v(q,\dot{q}))\]
Taking the partial derivatives with respect to the feature variable and
its rate, the components to the equation of motion are build the DAE we know from physics:
\[\frac{\mathrm{d}}{\mathrm{d}t}\frac{\partial L}{\partial \dot{q}} = \frac{\partial L}{\partial q}\].
By the chain rule, the velocities are related to the
feature space by
\begin{equation}
\dot{x} = \frac{\mathrm{d}x(q)}{\mathrm{d}t} = \frac{\partial x}{\partial q}\frac{\mathrm{d}q}{\mathrm{d}t}
\end{equation}
and similarly for the \(y\) direction.
Then plug it into a Runge Kutta (trapezoidal) to solve with Newton's
method:
\[\frac{\partial L}{\partial \dot{q}}_i - \Delta t \frac{\partial L}{\partial q}_i = \frac{\partial L}{\partial \dot{q}}_0 + \Delta t \frac{\partial L}{\partial q}_0\]

Soltutions are shown in Figure \ref{fig:pendulum}.
\begin{figure}
\centering
\includegraphics[width=5in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/pendulum_q.png}
\caption{\label{fig:pendulum}Solution of the pendulum on an arbitrary latent space.}
\end{figure}

@@

\section*{Acknowlegements}

This work was supported by Department of Energy, Office of Fossil
Energy contracts
1. Hydrates
2. Shales

The author would like to thank his colleagues Matt Reagan and George
Moridis for all their instruction on chemsitry. 
\bibliographystyle{plainnat}
\bibliography{zotero}
